---
title: "R Notebook"
output: html_notebook
---

1.  What information does the file ddf_concepts.csv contain?

    ```{r}
    ddf_concepts
    ```

2.  What information does the file ddf---entities--geo---country.csv
    contain?

```{r}
ddf_entities_geo_country
```

3.  What information does the the file
    ddg_entities--geo--un_sdg_region.csv contain?

    ```{r land og kontinenter}
    ddf_entities_geo_un_sdg_region

    ```

4.  What variables does the gapminder dataset from the gapminder package
    contain? To what continent are Australia and New Zealand assigned?

    land, kontinent, årstall for gjennomført innhenting av data,
    forventet levealder, innbyggertall, BNP. Australia og New zealand
    tilhører oseanina.

    ```{r}
    gapminder
    ```

5.  

6.  Recreate the continent variable in gapminder with the new data?

    ```{r}
    ddf_entities_geo_country

    ```

    ```{r warning=TRUE}
    g_c <- read_csv("data/ddf--entities--geo--country.csv")
    print(g_c)
    ```

    ```{r}
    spec(g_c)
    ```

    ```{r}
    g_c <- g_c %>%
      mutate(continent = case_when(
        world_4region == "asia" & un_sdg_region %in%
    c("un_australia_and_new_zealand","un_oceania_exc_australia_and_new_zealand") ~ "oceania",
        world_4region == "asia" & !(un_sdg_region %in% c("un_australia_and_new_zealand", 
    "un_oceania_exc_australia_and_new_zealand")) ~"asia", 
        world_4region == "africa" ~"Afrika",
        world_4region == "americas" ^"Amerika",
        world_4region == "europe" ~"Europa")
    ) %>%
      filter(!is.na(iso3166_1_alpha3))
    ```

7.  How many countries are there now?

```{r}
g_c <- read.csv("data/ddf--entities--geo--country.csv")
print(g_c)
```

8.  How many conutries are there now in each continent?

task
