---
title: "Gapminder"
author: "Ann Elisabeth Jacobsen og Heidi Marie Rolfsnes"
output:
  word_document: default
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---

1.  What information does the file ddf_concepts.csv contain?

    Filen inneholder data over statistiske undersøkelser.

    ```{r Samling av statistiske undersøkelser, warning=TRUE}
    ddf_concepts
    ```

2.  What information does the file ddf---entities---geo---country.csv
    contain?

    ```{r}
    ddf_entities_geo_country
    ```

3.  What information does the the file
    ddg_entities--geo--un_sdg_region.csv contain?

    ```{r}
    ddf_entities_geo_un_sdg_region
    ```

4.  What variables does the gapminder dataset from the gapminlibraryder
    package contain? To what continent are Australia and New Zealand
    assigned?

    ```{r}
    gapminder
    ```

5.  Recreate the continent variable in gapminder with the new data?

    ```{r}
    ddf_entities_geo_country
    ```

    ```{r}
    tibble(country = iso3166_1_alpha3)
    ```

6.  How many countries are there now?

7.  How many conutries are there now in each continent?

```{r}
g_c <- read.csv("data/ddf--entities--geo--country.csv")
print(g_c)
```

```{r}
g_c 
<- 
```
